Diagnosing Solver Failures
========================================================

## Runs included
We have a couple of test runs here.  They are:
- **nolog**: Mostly Broyden+linear mode alternating with the preconditioner.  The bisector gets to run only if there are markets that are being completely excluded from the Broyden solver. This version mostly failed to solve.
- **nlbf**:  Due to an oversight, the *nolog* run ran the bisector in between the preconditioner and the Broyden solver.  This version reverses the order, as intended, which causes periods 4-11 to solve.

## Notes on the raster plots
The raster plots display vector quantities as a function of iteration count, with iterations on the y-axis and components of the vector on the x-axis.  The number of markets being solved is variable, so unused vector components, if any, are colored black.

In some cases the data values have been scaled so as to provide better contrast on intervals of interest:
- The $F(x)$ raster plots have their values remapped to a partially logarithmic scale.  Values with magnitudes less than $10^{-3}$ are flushed to zero *(displayed in white)*.  (These markets would be considered solved under the default solution tolerance.)  Values between $10^{-3}$ and 1 are mapped logarithmically in the ranges -3-0 and 0-3, according to their sign *(yellow-green tints for values < 0, blue-green for values > 0)*.  Values between 1 and 10 are mapped linearly to the range 3-13 (-13 - -3, for negative values) *(red/blue tints)*, and values with magnitudes greater than 10 are clamped to +/- 13 *(saturated red/blue)*.
- Plots of changes by iteration ($\Delta x$ and $\Delta F(x)$) are shown as absolute values, with values > 10 clamped to 10.
- Plots of derivatives (whether partial or total) are clamped to the range [-100, 100].

The newer data inserts a one-pixel gap between calls to the Broyden solver.  Some of the older datasets lack this visual guide.  In either case, we see discontinuities whenever one call ends and the next begins.  These discontinuities are caused partly by the adjustments that other solvers, such as the preconditioner can make to prices; however, adding and removing markets from the solvable set also scrambles (slightly) the columns in the plots, even when the number of markets being solved is unchanged.

```{r setoptions, echo=FALSE, cache=FALSE}
opts_chunk$set(message=FALSE,dpi=100, cache=TRUE)
source("solver-diagnostics.R")
```

```{r readdata, echo=FALSE}
### Read several datasets of interest.
nolog.alldata <- read.trace.log("data/ctax-solver-nolog.txt")
nolog.per4 <- nolog.alldata$`4`
nolog.per4.fxvals  <- nolog.per4$fx
nolog.per4.dfxvals <- nolog.per4$deltafx
nolog.per4.dxvals  <- nolog.per4$deltax
nolog.per4.diagB   <- nolog.per4$diagB
nolog.per4.totalderiv <- calc.total.deriv(nolog.per4.dfxvals, nolog.per4.dxvals)

nlbf.alldata         <- read.trace.log("data/solver-data-nolog-broydenfirst.txt")
nlbf.per4            <- nlbf.alldata$`4`
nlbf.per4.fxvals     <- nlbf.per4$fx
nlbf.per4.dfxvals    <- nlbf.per4$deltafx
nlbf.per4.dxvals     <- nlbf.per4$deltax
nlbf.per4.diagB      <- nlbf.per4$diagB
nlbf.per4.totalderiv <- calc.total.deriv(nlbf.per4.dfxvals, nlbf.per4.dxvals)

## Also get period 12 for nlbf
nlbf.per12           <- nlbf.alldata$`12`
nlbf.per12.fxvals    <- nlbf.per12$fx
nlbf.per12.dfxvals   <- nlbf.per12$deltafx
nlbf.per12.dxvals    <- nlbf.per12$deltax
nlbf.per12.diagB     <- nlbf.per12$diagB
nlbf.per12.totalderiv <- calc.total.deriv(nlbf.per12.dfxvals, nlbf.per12.dxvals)
```

The plots below show period 4 of the nolog case, which failed to solve.  Discussion follows the plots.

```{r nolog.heatmaps4fx, fig.width=10, fig.height=8}
heatmap.fx(nolog.per4.fxvals, "F(x) values, run= nolog, per= 4")
```

Several patterns are apparent in this plot:

1. Although many markets appear to get close to solution, very few are actually solved.
2. For $|F(x)|>1$, insufficient demand (negative values / red tints) is much more common than excess demand.  This does not appear to be the case for $|F(x)|<1$.  
3. The reset that occurs when the solver stalls frequently results in the solver moving *further* from solution.  The block of iterations ending around iteration 320 is an especially vivid example.
4. Several markets seem to be persistently problematic.  Looking at the final block of iterations and cross-referencing with the solver log we can identify some of them:
  + **20--22**: GlobalRice, GlobalSugarCrop, GlobalWheat
  + **150--160**: Various regions "electricity-net-ownuseDemand-int" markets
  + **170--181**: Various regions FodderGrass, plus globalFodderHerb, globalMiscCrop, and globalRoot-Tuber

```{r nolog.heatmaps4, fig.width=10, fig.height=8}
heatmap.delta(nolog.per4.dxvals, "Abs(delta-x), run= nolog, per= 4")
heatmap.dfdx(nolog.per4.diagB, "Diagonal partials, run= nolog, per= 4")
heatmap.dfdx(nolog.per4.totalderiv, "Total derivatives, run= nolog, per= 4")
```

The $\Delta x$ plot shows that many of the vector components are limited to tiny steps at each iteration.  The changes in the price vector are dominated by a few markets.  In a couple of iteration blocks, such as the ones surrounding iterations 340 and 420, it appears that the solver algorithm made very little progress at all before bailing out.  The components in the block around 150--160 are demand markets.  As such their "prices" are actually demand values, with a different characteristic scale, so it is perhaps not surprising that their step values are larger.  However, for the rest of the components it looks as if the algorithm is stuck for much of its run and that most of the progress, such as it is, takes place between iteration blocks.  

```{r nlbf.heatmaps4, fig.width=10, fig.height=8}
heatmap.fx(nlbf.per4.fxvals, "F(x) values, run= nlbf, per=4")
heatmap.delta(nlbf.per4.dxvals, "Abs(delta-x), run= nlbf, per= 4")
heatmap.dfdx(nlbf.per4.diagB, "Diagonal partials, run= nlbf, per= 4")
heatmap.dfdx(nlbf.per4.totalderiv, "Total derivatives, run= nlbf, per=4")
```
